\subsection{SSRadar - Single Scattering Radar \index{SSRadar}}

A simple single scattering radar simulator which assumes pure Rayleigh scattering
in a 1D plan-parallel atmosphere at water or ice spheres without molecular absorption. 
It solves the equation for the radar reflectivity factor $Z$
$$
Z = \sum_i N_i D_i^6
$$

directly from the given droplet distribution defined by discrete sampling intervalls
$\left(N_i, N_i + \Delta N\right)$ containing the droplet diameters $\left(D_i, D_i 
+ \Delta D\right)$ (\citet{Rinehart2010}).

Usage (outputfile is optional, output is printed to stdout by default):
\begin{Verbatim}[fontsize=\footnotesize]
ssradar inputfile [outputfile]
\end{Verbatim}
The simulation is defined in the input file with the following structure:
\subsubsection {First line:}
\begin{description}
\item[wavelength] The wavelength in mm.
\item[Z-Position] The vertical position of the radar in m.
\item[Zenith Angle] The zenith angle in degree, $0$ means looking straight upwards, 
$180$ means looking straight downwards, must not be $90$.
\item[Ground Altitude] Vertical position of the ground in m.
\end{description}
Notes: The ground altitude can be used to add topographic data when creating two- or 
three-dimensional radar images with multiple simulations. The signal from the ground is
set to $100$ $dBZ$.
\subsubsection{Second line:}
\begin{description}
\item[Range Gates] Number of range gates.
\item[Range Gate Length] Radial Length of a range gate in m.
\item[First Range Gate] Radial Distance of the first range gate from the radar in m.
\item[Output] Ouput format. -1 is text only, 1 is mmclx only, 0 is both.
\end{description}
Notes: The simulator takes a straight radial line from the radar into the direction defined
by the zenith angle. This line is partitioned into a number of intervals (number of 
range gates) with a fixed length (range gate length) starting at a given distance 
(first range gate) from the radar. Keep in mind that the first range gate is defined by 
its distance along that line and not by its z-position, and that the length of a range
gate is the actual geometrical length. The distinction between first range gate and position
of the radar serves to be able to only focus on the parts that are needed and it will be
further important once attenuation has been implemented. MMCLX output has been added for
easy compatibility.
\subsubsection{The following lines define the reflective layers, one line per layer:}
\begin{description}
\item[Height] Z-Position of the lower border of the layer in m.
\item[Thickness] Vertical thickness of the layer in m.
\item[Effective Radius] Effective Radius of the distribution in um.
\item[Distribution] What kind of distribution (mono, gamma, log, or a filename), for
details see \code{distribution} and \code{size\_distribution\_file} in chapter \ref{sec:mie}.
\item[Distribution Parameter] Distribution Parameter for gamma and log, for
details see \code{distribution} in chapter \ref{sec:mie}.
\item[Maximum Radius Factor] Upper radius cutoff factor for gamma and log, for
details see \code{n\_r\_max} in chapter \ref{sec:mie}.
\item[Temperature] Temperature of the layer in degree Celsius.
\item[Phase] Phase of the layer, water or ice.
\item[Liquid Water Content] Liquid Water Content of the layer in $g/m^3$
\end{description}
Notes: You can simulate supercooled liquid water down to $-34^{\circ}C$.
Reflective layers must not overlap or be below ground but can otherwise be 
configured freely. 

\subsubsection{Example Input:}
\begin{Verbatim}[fontsize=\scriptsize,frame=single]
# wvl - z-pos -   zenith angle - ground altitude
# mm  - m     -   degree       - m
  8.5   500.0       45.0        0.0
# Range Gates - RG Length - First RG - Output
# (int)         m           m          (-1,0,1)
  15            500.0       0.0        -1
# height - thickness - reff - dist - distparam - max r fact - T - phase - LWC
# m        m           um     (str)  (double)    (double)  deg C  (str)   g/m^3
  1000.0   1000.0      18.0   mono       0          0        10.  water   0.30
  2000.0    500.0      15.0   log        1.1        5         0.  water   0.10
  3000.0    500.0      12.0   gamma      7          5       -10.  water   0.05
  3500.0   2000.0      10.0   dist.dat   0          0       -20.  ice     0.02
\end{Verbatim}

\subsubsection{Example Output:}
\begin{Verbatim}[fontsize=\scriptsize,frame=single]
Wavelength: 8500 um
Height of radar: 0.000000e+00 m
Zenith Angle: 0.000000e+00
Number of Range Gates: 6
Range Gate Width: 1.000000e+03 m
First Range Gate Radial Distance: 0.000000e+00 m
First Range Gate Z-Position: 0.000000e+00 m
RG No    RG Dist [m]     RG Z-Pos [m]    Z [dBZ]
1        0.000000e+00    0.000000e+00    0.000000
2        1.000000e+03    1.000000e+03    -16.016457
3        2.000000e+03    2.000000e+03    -19.641281
4        3.000000e+03    3.000000e+03    -23.466286
5        4.000000e+03    4.000000e+03    -27.732285
6        5.000000e+03    5.000000e+03    0.000000
\end{Verbatim}

